Can we distinguish modes of selective interactions using linkage disequilibrium?

### 2021-03-13 Manuscript.Rmd
asses to assess
edits and readme

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting patterns of allele frequencies and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over interpretation of
observations. Interpreting this data is hindered by the relative lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection, with arbitrary
epistasis and dominance, and variable recombination and population sizes. I use
this to explore how different models of epistasis and dominance affect expected
signed LD, including for non-steady-state demography relevant to human
populations. Finally, I use whole-genome sequencing data from humans to asses
how well we can differentiate modes of selective interactions in practice. I
find that positive LD between missense mutations within genes is driven by
strong positive correlations between pairs of mutations that fall within the
same conserved domain, pointing to compensatory mutations or antagonistic
epistasis as the prevailing mode of interaction within but not outside of
conserved genic elements. The heterogeneous landscape of both strength of
selection and selective interactions within protein-coding genes calls for more
refined inferences of the joint distribution of fitness and interactive
effects, and the methods presented here should prove useful in that pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus. Linked
selection reduces the fixation probability of advantageous mutations and
increases that of deleterious mutations compared to expectations under
single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
selected variants are also predicted to deviate from expectations without
interference. Under a multiplicative fitness model, where the fitness reduction
of a haplotype carrying multiple deleterious variants is equal to the product
of the fitness reduction of each mutation independently, we should expect to
see those mutation segregate on different haplotypes more often than together,
leading to negative, or repulsion, linkage disequilibrium (LD), although the
extent of LD depends non-trivially on the strength of selection and probability
of recombination separating loci [@Hill1966-gv; @McVean2000-ox].

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
synergistic epistasis, where multiple deleterious mutations had a greater
effect on fitness than the multiplication of the observed effects of the
individual mutations. The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems which can be compared to evolutionary models
[@Otwinowski2018-zk].

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and test inference methods, but do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution, or
two-locus haplotype frequency spectrum, stores the expected density or observed
counts of pairs of loci with each possible haplotype frequency configuration in
a sample of size $n$. In many cases it is simpler to visualize summaries (such
as $\E[D]$) or slices of the distribution (Figure \@ref(fig:fig1)) instead of
the full three-dimensional distribution. Here, I focus on low-order LD
statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
distance as these are statistics that are commonly used to test for
interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$. These statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

For additive mutations (i.e. $h=1/2$) and no epistasis, we recover the well
known Hill-Robertson interference result of negative LD between selected
mutations, which is strongest for pairs for mutations that have selection
coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure \@ref(fig:fig1)D).
For strongly deleterious mutations, LD is close to zero even with tight
linkage, as they almost always segregate at low enough frequencies that they
are unlikely to interfere with each other [@McVean2000-ox].

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
antagonistic epistasis results positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
considered moderate, large mean LD extends well beyond 100 kilobases, with more
strongly deleterious interacting mutations expected to show large signed LD
over much larger recombination distances.

@@

The effects of epistasis between variants on LD has been widely documented.
However, the effect of non-additive selection within a locus on interference
has received less attention. @Garcia2020-zn used large-scale forward simulation
to explore the effect of dominance on patterns of LD, showing that LD depends
on both the magnitude of the fitness and dominance effects in a nonlinear way,
but note that expensive large-scale simulations are needed to explore these
effects as no closed form approach to computing those expectations exist. The
method presented here provides such a closed form approach, so that
expectations of signed LD and other two-locus statistics can be computed more
efficiently and accurately.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

The case with $s_{AB/ab}\not=s_{Ab/aB}$ could arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid carrying at least one copy that is free of mutations has
minimal fitness loss. This is analogous to the single-locus dominance case,
where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig4)E-F to G-H), although the interpretation of those two
models can vary. It does show that with a highly parameterized space of
possible general diploid selection models, multiple models with different
biological interpretations can give similar patterns of expected $D$.

>>>>

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over the interpretation of
observations from data. Interpreting this data is hindered by the relative lack
of analytic or even numerical expectations for patterns of variation between
pairs of loci under the combined effects of selection, dominance, epistasis,
and demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive correlations between pairs of mutations that fall within the same
conserved domain, pointing to compensatory mutations or antagonistic epistasis
as the prevailing mode of interaction within but not outside of conserved genic
elements. The heterogeneous landscape of both mutational fitness effects and
selective interactions within protein-coding genes calls for more refined
inferences of the joint distribution of fitness and interactive effects, and
the methods presented here should prove useful in that pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus. Linked
selection reduces the fixation probability of advantageous mutations and
increases that of deleterious mutations compared to expectations under
single-locus models [@Birky1988-jm]. Allele frequency dynamics and correlations
of linked selected variants are also predicted to deviate from expectations
without interference. Under a multiplicative fitness model, where the fitness
reduction of a haplotype carrying multiple deleterious variants is equal to the
product of the fitness reduction of each mutation independently, we should
expect to see those mutation segregate on different haplotypes more often than
together, leading to negative, or repulsion, linkage disequilibrium (LD),
although the extent of LD depends non-trivially on the strength of selection
and probability of recombination separating loci [@Hill1966-gv;
@McVean2000-ox].

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
synergistic epistasis, where multiple deleterious mutations had a greater
effect on fitness than the multiplication of the observed effects of the
individual mutations. The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems which can be compared to evolutionary theory
[@Otwinowski2018-zk].

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their rarity and low allele frequency mean that statistical measurement of LD
and other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and test inference methods, but do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
the distribution (Figure \@ref(fig:fig1)) instead of the full three-dimensional
distribution. Here, I focus on low-order LD statistics, such as $\E[D]$ and
$\E[D^2]$, and their decay with recombination distance as these are statistics
that are commonly used to test for interactions between loci. Instead of
unnormalized $\E[D^2]$ and $\E[D]$, I consider expectations for $\sigma_d^2 =
\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
and $q$ are the derived allele frequencies at the left ($A/a$) and right
($B/b$) loci, respectively. These normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

For mutations under additive selection (i.e. $h=1/2$) and no epistasis, we
recover the well known Hill-Robertson interference result of negative LD
between selected mutations, which is strongest for pairs for mutations that
have selection coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure
\@ref(fig:fig1)D). For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other [@McVean2000-ox].

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig2)D even moderately deleterious mutations with
population-size-scaled selection coefficients of $\gamma=-10$ show large mean
LD that extends well beyond 100 kilobases. More strongly deleterious
interacting mutations are expected to show large signed LD over much larger
recombination distances.

@@

The effects of epistasis between variants on LD have been widely studied.
However, the combined effect of non-additive selection within a locus and
interference has received less attention. @Garcia2020-zn used large-scale
forward simulations to explore the effect of dominance on patterns of LD,
showing that LD depends on both the magnitude of the selection and dominance
coefficients in a nonlinear way, but note that expensive large-scale
simulations are needed to explore these effects as no closed form approach to
computing those expectations exist. The method presented here provides such a
closed form approach, so that expectations of signed LD and other two-locus
statistics can be computed more efficiently and accurately.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho \gg 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. This is analogous to the single-locus
dominance case, where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based
dominance" scenario [e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher
fitness than an $Ab/aB$ type (a simple implementation of this model is given in
the final column in Table \@ref(tab:selmodels)). Such a gene-based fitness
model gives similar expected positive signed LD to the model of antagonistic
epistasis (compare Figure \@ref(fig:fig4)E-F to G-H), although the
interpretation of those two models can differ. It shows that with a highly
parameterized space of possible general diploid selection models, multiple
models with different biological interpretations can give similar patterns of
expected $D$.
